---
title: "scRNAseq Analysis — {{PROJECT_NAME}}"
subtitle: "{{PROJECT_NAME}}"
author: "{{AUTHOR}}"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Table of contents"
    toc-expand: false
    number-sections: true
    colorlinks: true
    self-contained: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    smooth-scroll: true
    page-layout: full
    highlight-style: github
    theme: cosmo
    df-print: paged
    lightbox: true
    embed-resources: true
    fig-width: 12
    fig-height: 7
    fig-dpi: 150
include-in-header: styles/pfgt_style_light.html
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    comment: ""
    dev: "ragg_png"
---

```{r setup, include=FALSE}
library(scReportR)
library(patchwork)
library(DT)

cfg             <- sc_config("params.yml")
thresholds_list <- sc_qc_all_thresholds(cfg)
```

# Project overview

```{r overview}
tibble::tibble(
  Parameter = c("Date", "Samples", "Normalisation",
                "Integration", "Default resolution"),
  Value = c(
    as.character(Sys.Date()),
    nrow(sc_sample_meta(cfg)),
    cfg$normalization$method,
    cfg$integration$method,
    cfg$clustering$default_resolution
  )
) |> DT::datatable(rownames = FALSE, colnames = NULL,
                   options = list(dom = "t"))
```

```{r sample-table}
sc_sample_meta(cfg) |>
  dplyr::mutate(data_exists = file.exists(path)) |> dplyr::select(-path, -raw_path) |>
  DT::datatable(rownames = FALSE, caption = "Sample table",
                options = list(scrollX = TRUE))
```

---

# Quality Control

::: callout-note
Quality control identifies and removes low-quality cells before downstream
analysis. Key metrics are: number of detected genes (`nFeature_RNA`),
total UMI counts (`nCount_RNA`), percentage of mitochondrial reads
(`percent.mt`), and cell complexity (`log10(nGenes)/log10(nUMI)`).

High mitochondrial content typically indicates a damaged or dying cell.
Low complexity suggests an empty droplet or poor-quality cell.
:::

## Cell counts

```{r load-data}
obj_list <- sc_load(cfg)
```

```{r cell-counts, fig.height=4}
sc_plot_cell_counts(obj_list, x = "sample_id", fill = "library_type",
                    cfg = cfg, title = "Raw cell counts per sample")
```

## Genes and UMI distribution

```{r gene-counts, fig.height=4}
sc_plot_qc_distri(obj_list, cfg = cfg)
```

::: callout-note
Each point is one sample. Wide spread across samples may indicate
variable sequencing depth or technical differences between library
types — check the violin plots below for within-sample distributions.
:::

::: {.callout-note collapse="true"}
## Summary table

```{r qc-summary-table}
sc_qc_summary(obj_list) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```
:::

## Configured thresholds

```{r thresholds-plot, fig.height=3.5}
sc_plot_thresholds(thresholds_list, cfg)
```

::: {.callout-tip collapse="true"}
## Full threshold table

```{r thresholds-table}
sc_qc_thresholds_df(thresholds_list, cfg) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

Adjust per-sample overrides in `params.yml → qc → per_sample`, then
re-render. Red items above have a per-sample override active.
:::

## QC distributions

```{r qc-violins, fig.height=6}
sc_plot_qc_violins(obj_list, thresholds_list, cfg)
```

::: callout-note
Red dashed lines show per-sample thresholds. Segments at different
heights indicate per-sample overrides configured in `params.yml`.
:::

## Scatter plots {.tabset}

```{r qc-scatter, results='asis', fig.height=4}
sc_plot_qc_scatter(obj_list, thresholds_list) |>
  sc_tabset()
```

## Cell complexity

```{r complexity, fig.height=6}
sc_plot_complexity(obj_list, cfg)
```

::: callout-note
Complexity (`log10(nGenes) / log10(nUMI)`) identifies degraded cells
independently of absolute gene count thresholds. Cells below 0.8
(dashed line) are typically removed — they often represent empty
droplets or cells with abnormally few genes relative to their UMI count.
:::

---

# Preprocessing

::: callout-note
Preprocessing removes technical artefacts before biological analysis:

1. **Permissive pre-filter** — removes obvious empty droplets
   (nFeature < 100) without biasing ambient RNA estimation.
2. **Ambient RNA decontamination** (decontX) — estimates and removes
   RNA from lysed cells that contaminates the droplet soup.
3. **Doublet detection** (scDblFinder) — flags droplets containing
   two cells, which would otherwise appear as artificial hybrid cell types.
4. **Final QC filtering** — applies the per-sample thresholds above.
:::

## Ambient RNA decontamination (decontX)

```{r decontX}
obj_list_prefilt <- lapply(obj_list, sc_prefilter)
obj_list_decontX <- sc_decontX(obj_list_prefilt, cfg)
```

```{r decontX-status}
tibble::tibble(
  Sample     = names(obj_list_decontX),
  Background = sapply(obj_list_decontX, function(o) {
    ifelse(isTRUE(unique(o$decontX_background)),
           "\u2714 raw matrix", "\u26a0 no background")
  })
) |> DT::datatable(rownames = FALSE, options = list(dom = "t"))
```

```{r decontX-plot, fig.height=5}
sc_plot_decontX(obj_list_decontX, cfg)
```

::: callout-note
- Score near **0** = clean cell
- Score near **1** = heavily contaminated (ambient RNA)

Samples without a `raw_path` in `params.yml` ran in no-background mode
(⚠) — scores are less accurate. Add `raw_path` pointing to
`raw_feature_bc_matrix` to improve estimation.

The `decontX$use_decontX_counts` option in `params.yml` (default: `false`)
replaces raw counts with decontaminated counts for all downstream steps.
:::

## Doublet detection (scDblFinder)

```{r doublets}
obj_list_dbl <- sc_doublets(obj_list_decontX, cfg)
```

```{r doublets-plot, fig.height=4}
sc_plot_doublets(obj_list_dbl, cfg)
```

::: callout-tip
Expected doublet rate for 10x Genomics is approximately 0.8% per 1,000
cells loaded. Higher rates may indicate overloading of the chip.
Doublets are removed in the filtering step below.
:::

## Filtering

```{r filter}
obj_list_filtered <- sc_filter(obj_list_dbl, thresholds_list,
                                remove_doublets = TRUE)
```

```{r filter-waterfall, fig.height=4}
waterfall <- dplyr::bind_rows(
  tibble::tibble(sample  = names(obj_list),
                 step    = "1. Raw",
                 n_cells = unname(sapply(obj_list, ncol))),
  tibble::tibble(sample  = names(obj_list_prefilt),
                 step    = "2. Pre-filter",
                 n_cells = unname(sapply(obj_list_prefilt, ncol))),
  tibble::tibble(sample  = names(obj_list_filtered),
                 step    = "3. Final",
                 n_cells = unname(sapply(obj_list_filtered, ncol)))
) |>
  dplyr::mutate(step = factor(step,
                              levels = c("1. Raw", "2. Pre-filter", "3. Final")))

sc_plot_filter_waterfall(waterfall, cfg)
```

::: {.callout-note collapse="true"}
## Filtering summary table

```{r filter-table}
tibble::tibble(
  Sample  = names(obj_list),
  Raw     = sapply(obj_list, ncol),
  Final   = sapply(obj_list_filtered[names(obj_list)], ncol),
  Removed = Raw - Final,
  Pct     = round(100 * Removed / Raw, 1)
) |> DT::datatable(rownames = FALSE)
```
:::

## Merge

```{r merge}
merged <- sc_merge(obj_list_filtered, cfg)
```

```{r merge-composition, fig.height=4}
# Adjust x and fill to the metadata columns available in your project
sc_plot_cell_counts(merged, x = "sample_id", fill = "library_type",
                    cfg = cfg, title = "Merged object composition")
```

::: callout-note
In Seurat v5, RNA layers remain **separate** after merge — this is
intentional. `JoinLayers()` is called automatically after integration
to enable `FindMarkers()`.
:::

---

# Normalisation & Integration

::: callout-note
**Normalisation** (`r cfg$normalization$method`) removes technical
variation in sequencing depth so that gene expression values are
comparable across cells.

**Integration** (`r cfg$integration$method`) corrects for batch effects
between samples, aligning cells of the same type regardless of which
sample they came from, while preserving true biological variation.
:::

## Normalisation

```{r normalize}
merged <- sc_normalize(merged, cfg)
```

## Highly variable genes

```{r hvg, fig.height=5}
sc_plot_hvg(merged, n_label = 20)

```

::: callout-note
Highly variable genes (HVGs) capture the most biologically informative
variation across cells and are used as input for PCA. The top 20 are
labelled above.
:::

## PCA

```{r pca-tabset, results='asis', fig.height=6, fig.width=10}
# Show cells coloured by key metadata variables in PCA space
# (before integration — useful to visualise batch effects)
sc_plot_pca_cols(merged,
  cols = c("sample_id", "library_type"),
  cfg  = cfg
) |> sc_tabset()
```

```{r elbow, fig.height=3, fig.width=4}
sc_plot_elbow(merged, cfg)
```

::: callout-tip
Use the elbow plot to choose `pcs_use` in `params.yml`. The red line
shows the current value (`r cfg$normalization$pcs_use` PCs). Pick the
point where the curve flattens — adding more PCs provides diminishing
returns.
:::

## Integration

Corrected covariates: `r paste(unlist(cfg$integration$group_by), collapse = ", ")`

```{r integrate}
merged <- sc_integrate(merged, cfg)
```

```{r integration-check, results='asis', fig.height=6, fig.width=10}
sc_plot_umap_cols(merged,
  cols = c("sample_id", "library_type"),
  cfg  = cfg
) |> sc_tabset()
```

::: callout-note
After successful integration, cells of the same type should overlap
across samples while retaining biological separation. If samples remain
clustered separately, try increasing `harmony_theta` (stronger
correction) or switching to `RPCAIntegration` in `params.yml`.
:::

---

# Clustering

::: callout-note
Clustering groups cells with similar transcriptional profiles. The
**Leiden algorithm** (algorithm 4) is used here — it produces more
stable communities than Louvain for single-cell data.

Multiple resolutions are tested simultaneously. Higher resolutions
produce more, finer-grained clusters. Use **clustree** to identify the
resolution where cluster assignments stabilise, then set
`default_resolution` in `params.yml`.
:::

Algorithm: **`r ifelse(cfg$clustering$algorithm == 4, "Leiden", "Louvain")`**
| Resolutions: `r paste(cfg$clustering$resolutions, collapse = ", ")`

```{r cluster}
merged <- sc_cluster(merged, cfg)
```

## Cluster stability (clustree)

```{r clustree, out.width='70%'}
library(clustree)
sc_plot_clustree(merged, cfg$clustering$resolutions)
```

::: callout-tip
In clustree, stable clusters appear as nodes that persist across multiple
resolutions. Nodes that immediately split may indicate over-clustering.
Choose the resolution just before major instability appears.
:::

## UMAP — all resolutions {.tabset}

```{r umap-resolutions, results='asis', fig.height=6, fig.width=10}
sc_tabset_resolutions(merged, cfg$clustering$resolutions)
```

## UMAP — default resolution (`r cfg$clustering$default_resolution`)

```{r umap-clustering, results='asis', fig.height=6, fig.width=10}
res_col <- paste0("clusters_res", cfg$clustering$default_resolution)

sc_plot_umap_cols(merged,
  cols = c(res_col, "sample_id", "library_type"),
  cfg  = cfg
) |> sc_tabset()
```

## Cluster composition

```{r cluster-composition, results='asis', fig.height=5}
# Adjust cols to the metadata columns available in your project
sc_plot_cluster_composition(merged,
  cols = intersect(c("library_type", "tissue", "population"),
                   colnames(merged@meta.data)),
  cfg  = cfg
)
```

::: callout-note
Clusters dominated by a single sample or library type may reflect
technical rather than biological differences. These should be examined
carefully before annotation.
:::


## Meta composition

```{r meta-composition, results='asis', fig.height=5}
sc_plot_cluster_composition(merged,
  cols    = c("sample_id","tissue", "library_type", "population"),
  cfg     = cfg,
  reverse = TRUE, legend="right"
)
```

---

# Marker Analysis

::: callout-note
Differential expression between clusters identifies the genes that best
define each cluster (**top markers**). These are used to:

1. Build a **heatmap** summarising cluster identity at a glance.
2. Compute **module scores** — a per-cell signature score for each
   cluster's marker programme, projected on the UMAP.
3. Cross-reference against **known canonical markers** to guide
   manual annotation.
:::

## Top markers per cluster

```{r find-markers}
# Remove unannotated genes and other elements to facilitate annotation
markers_df <- sc_find_markers(merged, cfg) |>
  dplyr::filter(
    !grepl("^ENSG|^LINC|^AC[0-9]|^AL[0-9]", gene),
    pct.1 >= 0.2    # au moins 20% des cellules expriment le gène
  )
```


```{r heatmap-callout, results='asis'}
if ("SCT" %in% names(merged@assays)) {
  cat("::: callout-note\n")
  cat("The heatmap uses scaled RNA counts (`RNA` assay), not SCT residuals.\n")
  cat("This ensures all marker genes can be displayed regardless of whether\n")
  cat("they are in SCT `scale.data`. Gene ordering and colour scale reflect\n")
  cat("log-normalised, z-scored expression.\n")
  cat(":::\n")
}
```

```{r markers-heatmap, fig.height=12}
sc_plot_markers_heatmap(merged, markers_df, n_top = 5)
```

::: {.callout-note collapse="true"}
## Full markers table

```{r markers-table}
if (!is.null(markers_df) && nrow(markers_df) > 0) {
  markers_df |>
    dplyr::group_by(cluster) |>
    dplyr::slice_max(avg_log2FC, n = 100, with_ties = FALSE) |>
    DT::datatable(rownames = FALSE, filter = "top",
                  options  = list(scrollX = TRUE, pageLength = 20))
}
```
:::

## Cluster module scores

```{r module-scores}
merged <- sc_module_scores(merged, markers_df, cfg, n_top = 20)
```

```{r module-scores-umap, results='asis', fig.height=6}
sc_plot_module_scores(merged, cfg)
```

::: callout-note
Each panel shows the module score for one cluster's top `r 20` marker
genes projected onto the UMAP. Cells of the corresponding cluster
should show the highest scores (warm colours). Diffuse high scores in
multiple clusters may indicate shared cell programmes (e.g. cycling
cells, stress response).
:::

## Known marker projection {.tabset}

```{r known-markers, results='asis', fig.height=6}
cat("\n::: {.panel-tabset}\n")
for (cell_type in names(cfg$markers)) {
  cat("\n##", cell_type, "\n\n")
  sc_plot_umap_cols(merged,
                    cols = unlist(cfg$markers[[cell_type]])) |>
    sc_tabset()
  cat("\n")
}
cat("\n:::\n")
```

## DotPlot — global view

```{r dotplot, fig.height=8, fig.width=16}
sc_plot_markers_dot(merged, cfg, markers_df = markers_df, n_top = 5)
```

::: callout-note
The dot plot shows average expression (colour) and percentage of
expressing cells (size) for each marker gene across clusters.
Canonical markers expressed in a cluster confirm the cell type identity
suggested by automated annotation below.
:::

---

# Cell Type Annotation

::: callout-note
Automated annotation uses reference datasets to assign cell type labels
to each cluster without manual inspection of every marker gene.

**SingleR** compares each cell's expression profile to bulk RNA-seq
reference datasets (celldex) and assigns the most similar cell type
label.

**Pan-Human Azimuth** uses a neural network classifier trained on 23
human tissues to assign hierarchical cell ontology labels.

Automated annotations should always be validated against the known
marker projections above before being used in downstream analyses.
:::

## SingleR

```{r singler}
merged <- sc_singler(merged, cfg)
```

```{r singler-umaps, results='asis', fig.height=5}
if (isTRUE(cfg$annotation$singler$run)) {
  cat("\n::: {.panel-tabset}\n")
  for (ref in cfg$annotation$singler$references) {
    cat("\n##", ref, "\n\n")
    sc_plot_umap_cols(merged,
      cols = c(
        paste0("singler_", gsub("Data$", "", ref), "_label"),
        paste0("singler_", gsub("Data$", "", ref), "_pruned")
      )
    ) |> sc_tabset()
    cat("\n")
  }
  cat("\n:::\n")
}
```

```{r singler-heatmaps, results='asis', fig.height=7}
if (isTRUE(cfg$annotation$singler$run)) {
  cat("\n::: {.panel-tabset}\n")
  for (ref in cfg$annotation$singler$references) {
    cat("\n##", ref, "\n\n")
    print(sc_plot_singler_heatmap(merged, ref, cfg))
    cat("\n")
  }
  cat("\n:::\n")
}
```

::: callout-note
**Pruned labels** (second tab per reference) set uncertain predictions
to `NA` — cells where no reference type is a good match. High
proportions of `NA` cells may indicate a novel cell type, low quality,
or a mismatch between your tissue and the reference.
:::

## Pan-Human Azimuth

```{r panhuman-azimuth}
merged <- sc_panhuman_azimuth(merged, cfg)
```

```{r azimuth-umaps, results='asis', fig.height=6, eval=isTRUE(cfg$annotation$panhuman_azimuth$run)}
az_cols <- c("azimuth_broad", "azimuth_medium", "azimuth_fine")
az_cols <- az_cols[az_cols %in% colnames(merged@meta.data)]
sc_plot_umap_cols(merged, cols = az_cols) |> sc_tabset()
```

```{r azimuth-score, fig.height=5, eval=isTRUE(cfg$annotation$panhuman_azimuth$run)}
if ("final_level_softmax_prob" %in% colnames(merged@meta.data)) {
  sc_plot_umap_cols(merged, cols = "final_level_softmax_prob")[[1]]
}
```

::: callout-note
`azimuth_broad` → `azimuth_medium` → `azimuth_fine` represent
increasingly granular levels of the cell type hierarchy.
`final_level_softmax_prob` is the prediction confidence — cells below
0.5 should be interpreted with caution.
:::

## Manual annotation

::: callout-tip
Fill in the vector below using the marker heatmap, known marker
projections, and automated annotation results as guides. Cluster IDs
are listed in the suggestion table below.
:::

```{r singler-suggestion}
ref1 <- cfg$annotation$singler$references[[1]]
if (!is.null(ref1)) {
  col_lab     <- paste0("singler_", gsub("Data$", "", ref1), "_label")
  cluster_col <- paste0("clusters_res", cfg$clustering$default_resolution)

  if (col_lab %in% colnames(merged@meta.data)) {
    merged@meta.data |>
      dplyr::group_by(.data[[cluster_col]]) |>
      dplyr::count(.data[[col_lab]]) |>
      dplyr::slice_max(n, n = 1) |>
      dplyr::rename(Cluster = 1, Top_SingleR = 2, N_cells = n) |>
      DT::datatable(rownames = FALSE,
                    caption  = paste0("Top SingleR label per cluster (",
                                      ref1, ")"))
  }
}
```

```{r manual-annotation}
# ⚠️  Fill in after inspecting annotation results above  ⚠️
#
# annotation_manual <- c(
#   "1"  = "CD8 T cells",
#   "2"  = "CD4 T cells",
#   "3"  = "NK cells",
#   "4"  = "B cells",
#   "5"  = "Monocytes",
#   "6"  = "Tumor cells",
#   "7"  = "Fibroblasts",
#   "8"  = "Treg",
#   "9"  = "pDC",
#   "10" = "Plasma cells",
#   "11" = "Cycling"
# )
#
# merged <- sc_annotate_manual(merged, annotation_manual, cfg)
# sc_plot_annotation(merged)
```

---

# Export

```{r export}
sc_export_metadata(merged, cfg)

qs2::qs_save(merged,
             file.path(sc_cache_dir(cfg), "final_annotated.qs2"))
```

::: callout-note
To free disk space after finalising the analysis, run:

```r
sc_cache_finalize(cfg, dry_run = TRUE)   # preview what will be removed
sc_cache_finalize(cfg, dry_run = FALSE)  # confirm deletion
```

This keeps only `01_raw_objects.qs2` and `final_annotated.qs2`.
The full analysis can be reproduced from the raw object at any time.
:::
